---
title: "Assignment 5: Data Visualization"
author: "Meg O'Brien"
date: "Fall 2025"
output: pdf_document
geometry: margin=2.54cm
fig_height: 3
fig_width: 4
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data 
Visualization 

---

## Set up your session 

1. Set up your session. Load the tidyverse, here & cowplot packages, and verify
your home directory. Read in the NTL-LTER processed data files for nutrients and chemistry/physics for Peter and Paul Lakes (use the tidy `NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv` version in the 
Processed_KEY folder) and the processed data file for the Niwot Ridge litter 
dataset (use the `NEON_NIWO_Litter_mass_trap_Processed.csv` version, again 
from the Processed_KEY folder). 

2. Make sure R is reading dates as date format; if not change the format to 
date.

```{r initialize project}
#1 

library(tidyverse)
library(here)
library(ggplot2)
library(cowplot)
library(lubridate)
library(viridis)
library(RColorBrewer)
library(colormap)

getwd()

NTL_LTER <- read.csv(
  file = here
('Data/Processed_KEY/NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv')
, stringsAsFactors = TRUE)

NEON_NIWO <- read.csv(
  file = here('Data/Processed_KEY/NEON_NIWO_Litter_mass_trap_Processed.csv'), stringsAsFactors = TRUE)

#2 

class(NTL_LTER$sampledate)
class(NEON_NIWO$collectDate)

#Changing factor classes to dates

NTL_LTER$sampledate <- as.Date(NTL_LTER$sampledate, format = "%Y-%m-%d")
NEON_NIWO$collectDate <- as.Date(NEON_NIWO$collectDate, format = "%Y-%m-%d")

```
## Define your theme
3. Build a theme and set it as your default theme. Customize the look of at 
least two of the following: 
 * Plot background
 * Plot title
 * Axis labels
 * Axis ticks/gridlines
 * Legend

```{r build plot theme}
#3

mytheme <- theme_cowplot()+
  theme(
    plot.background = element_rect(fill = "mistyrose", color = NA),
    axis.text = element_text(color = "black"), 
    legend.position = "top")

theme_set(mytheme)

```

## Create graphs
For numbers 4-7, create ggplot graphs and adjust aesthetics to follow best 
practices for data visualization. Ensure your theme, color palettes, axes, and 
additional aesthetics are edited accordingly.

4. [NTL-LTER] Plot total phosphorus (`tp_ug`) by phosphate (`po4`), with 
separate aesthetics for Peter and Paul lakes. Add line(s) of best fit using 
the `lm` method. Adjust your axes to hide extreme values (hint: change the 
limits using `xlim()` and/or `ylim()`).

```{r plot total P vs PO4}
#4 

P_vs_PO4 <- ggplot(NTL_LTER, aes(x = po4, y = tp_ug, color = lakename, 
                                 shape = lakename))+
  geom_point(size = 3, alpha = 0.8) +                               
  geom_smooth(method=lm,se=FALSE) +            
  labs(
    title = "Total Phosphorus vs. Phosphate",
    x = "Phosphate",
    y = "Total Phosphorus",
    color = "Lake Name",
    shape = "Lake Name"
  ) +
  xlim(0, 100) +        
  ylim(0, 200)
print(P_vs_PO4)

```


5. [NTL-LTER] Make three separate boxplots of (a) temperature, (b) TP, and 
(c) TN, with month as the x axis and lake as a color aesthetic. Then, create 
a cowplot that combines the three graphs. Make sure that only one legend is 
present and that graph axes are aligned. 

Tips:
* Recall the discussion on factors in the lab section as it may be helpful here. 
* Setting an axis title in your theme to `element_blank()` removes the axis 
title (useful when multiple, aligned plots use the same axis values)
* Setting a legend's position to "none" will remove the legend from a plot.
* Individual plots can have different sizes when combined using `cowplot`.

```{r Create boxplots}
#5 

#a) Temperature Plot
Temp_plot <- ggplot(NTL_LTER, aes(x = month, y = temperature_C, 
                                  color = lakename))+
  geom_boxplot()+
  scale_y_continuous(expand = expansion(mult = c(0.05, 0.1))) +
  labs(y = "Temperature (°C)", x = "Month") +
  mytheme +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "top", 
    axis.title.y = element_text(size = 10), 
    axis.text.y = element_text(size = 8)    
  )
print(Temp_plot)

#b) TP Plot 

TP_plot <- ggplot(NTL_LTER, aes(x = month, y = tp_ug, color = lakename))+
  geom_boxplot()+
  scale_y_continuous(expand = expansion(mult = c(0.05, 0.1))) +
  labs(y = "Total Phosphorus", x = "Month") +
  mytheme +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "none",
    axis.title.y = element_text(size = 10), 
    axis.text.y = element_text(size = 8)   
  )
print(TP_plot)

#c) TN Plot 

TN_plot <- ggplot(NTL_LTER, aes(x = month, y = tn_ug, color = lakename))+
  geom_boxplot()+
  scale_y_continuous(expand = expansion(mult = c(0.05, 0.1))) +
  labs(y = "Total Nitrogen", x = "Month") +
  mytheme +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "none",
    axis.title.y = element_text(size = 10),  
    axis.text.y = element_text(size = 8)   
  )
print(TN_plot)

#Combine Plots 

legend <- get_legend(Temp_plot)

combined <- plot_grid(
  Temp_plot, TP_plot, TN_plot,
  ncol = 1,          
  align = "v"        
)

final_combined <- plot_grid(combined, legend, ncol = 1, rel_heights = c(1, 0.1))
print(final_combined)

```

Question: What do you observe about the variables of interest over seasons 
and between lakes? 

> Answer: Based on the observed boxplots of Peter and Paul Lake for temperature,
total phosphorus and total nitrogen over the sampling months, it is clear that 
Paul and Peter Lakes were sampled within different months of the year. Paul Lake
was sampled during the firt half of the year and Peter Lake was sampled during 
the second half of the year. Also, it is evident that Peter and Paul Lake have 
comprable temperatures, with Peter Lake showing slightly greater temperature 
variation. Peter Lake also shows higher total nitrogen and total phosphorus 
levels than Paul Lake. 

6. [Niwot Ridge] Plot a subset of the litter dataset by displaying only the 
"Needles" functional group. Plot the dry mass of needle litter by date and 
separate by NLCD class with a color aesthetic. (no need to adjust the name of 
each land use)

7. [Niwot Ridge] Now, plot the same plot but with NLCD classes separated into 
three facets rather than separated by color. 

```{r Plot litter}
#6

Needles_subset <- NEON_NIWO %>% 
  filter(functionalGroup == "Needles")

Litter_Needles <- ggplot(Needles_subset, aes(x = collectDate, y = dryMass, 
                                             color = nlcdClass)) +
  geom_line() +
  labs(
    x = "Date",
    y = "Dry Mass of Needles",
    color = "NLCD Class",
    title = "Needle Litter Dry Mass Over Time by NLCD Class"
  )
print(Litter_Needles)

#7

Litter_Needles_Faceted <- ggplot(Needles_subset, aes(x = collectDate, 
                                                     y = dryMass)) +
  geom_line() +
  labs(
    x = "Date",
    y = "Dry Mass of Needles",
    color = "NLCD Class",
    title = "Needle Litter Dry Mass Over Time by NLCD Class"
  ) +
  facet_wrap(vars(nlcdClass), nrow = 3)
print(Litter_Needles_Faceted)

```
Question: Which of these plots (6 vs. 7) do you think is more effective, and 
why? 

> Answer: I think the faceted plot is a better visualization because you can 
more clearly see the pattern of needle dry mass over the years in the three 
different NLCD Classes. The non-faceted plot was more difficult to analyze 
because the overlapping lines made it hard to read. 