---
title: "Assignment 3: Data Exploration"
author: "Meg O'Brien"
date: "Fall 2025"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Exploration.

---

## Set up your R session

1.  Load necessary packages (tidyverse, lubridate, here), check your current working directory and upload two datasets: the ECOTOX neonicotinoid dataset (ECOTOX_Neonicotinoids_Insects_raw.csv) and the Niwot Ridge NEON dataset for litter and woody debris (NEON_NIWO_Litter_massdata_2018-08_raw.csv). Name these datasets "Neonics" and "Litter", respectively. Be sure to include the subcommand to read strings in as factors.

```{r setup,  message = FALSE}

#install.packages('tidyverse')
#install.packages('lubridate')
#install.packages('here')

library(tidyverse)
library(lubridate)
library(here)

getwd()

Neonics <- read.csv(
  file = here('Data/Raw/ECOTOX_Neonicotinoids_Insects_raw.csv'), 
  stringsAsFactors = TRUE)

Litter <- read.csv(
  file = here('Data/Raw/NEON_NIWO_Litter_massdata_2018-08_raw.csv'), 
  stringsAsFactors = TRUE)


```

## Learn about your system

2.  The neonicotinoid dataset was collected from the Environmental Protection Agency's ECOTOX Knowledgebase, a database for ecotoxicology research. Neonicotinoids are a class of insecticides used widely in agriculture. The dataset that has been pulled includes all studies published on insects. Why might we be interested in the ecotoxicology of neonicotinoids on insects? Feel free to do a brief internet search if you feel you need more background information.

> Answer: Since neonicotinoids are one of the most commonly used insecticide in agriculture, it is important to understand their ecotoxicology to know how they will affect other organisms in the ecosystems in which they are introduced. For instance, neonicotinoids are highly toxic to bees, and their decline can trigger negative trophic cascades that can disrupt overall ecosystem balance. 

3.  The Niwot Ridge litter and woody debris dataset was collected from the National Ecological Observatory Network, which collectively includes 81 aquatic and terrestrial sites across 20 ecoclimatic domains. 32 of these sites sample forest litter and woody debris, and we will focus on the Niwot Ridge long-term ecological research (LTER) station in Colorado. Why might we be interested in studying litter and woody debris that falls to the ground in forests? Feel free to do a brief internet search if you feel you need more background information.

> Answer: Litter and woody debris are an essential component of forest ecosystems because they play a crucial role in nutrient cycling and decomposition processes. These functions help the ecosystem to maintain a healthy state. 

4.  How is litter and woody debris sampled as part of the NEON network? Read the NEON_Litterfall_UserGuide.pdf document to learn more. List three pieces of salient information about the sampling methods here:

> Answer: 
 1. Litter is is collected in elevated 0.5m^2 PVC traps.
 2. Woody debris is collected in ground traps. 
 3. Litter and woody debris are sampled at NEON sites that contain woody vegetation >2m tall. 

## Obtain basic summaries of your data (Neonics)

5.  What are the dimensions of the dataset?

```{r Q5}

dim(Neonics
    )

#The dataset has 4,623 rows and 30 columns. 
```

6.  Using the `summary` function on the "Effect" column, determine the most common effects that are studied. Why might these effects specifically be of interest? [Tip: The `sort()` command is useful for listing the values in order of magnitude...]

```{r Q6}

effect_counts <- table(Neonics$Effect)
sorted_effects <- sort(effect_counts, decreasing = TRUE)
sorted_effects
head(sorted_effects, 3)

```

> Answer: The three most common effects that are studied are population, mortality, and behavior. These are critical elements of population dynamics, and can be indicative of either positive or negative effects of insecticides on the population of an ecosystem. 


7.  Using the `summary` function, determine the six most commonly studied species in the dataset (common name). What do these species have in common, and why might they be of interest over other insects? Feel free to do a brief internet search for more information if needed.[TIP: Explore the help on the `summary()` function, in particular the `maxsum` argument...]

```{r Q7}

species.common.name_counts <- table(Neonics$Species.Common.Name)
sorted_species.common.name <- sort(species.common.name_counts, decreasing = TRUE)
sorted_species.common.name

```

> Answer: The six most commonly studied species in the dataset are the honey bee, parasitic wasp, buff tailed bumblebee, carniolan honey bee, bumble bee, and italian honeybee. These are all bee species; these are likely of interest over other species because of the negative effect of neonicotinoids on bees.

8.  Concentrations are always a numeric value. What is the class of `Conc.1..Author.` column in the dataset, and why is it not numeric? [Tip: Viewing the dataframe may be helpful...]

```{r Q8}

class(Neonics$Conc.1..Author.)

```

> Answer: The Conc.1..Author column is a factor class. This is because some of the values in this column contain less than or greater than symbols, which is a non-numeric character; this makes the column a factor instead of a numeric class. 

## Explore your data graphically (Neonics)

9.  Using `geom_freqpoly`, generate a plot of the number of studies conducted by publication year.

```{r Q9}

library(ggplot2)

ggplot(Neonics, aes(x = Publication.Year)) +
  geom_freqpoly(binwidth = 1) +
  labs(
    title = "Number of Studies by Publication Year",
    x = "Publication Year",
    y = "Number of Studies"
  ) 

```

10. Reproduce the same graph but now add a color aesthetic so that different Test.Location are displayed as different colors.

```{r Q10}

ggplot(Neonics, aes(x = Publication.Year, color = Test.Location)) +
  geom_freqpoly(binwidth = 1, size = 1.2) +
  labs(
    title = "Number of Studies by Publication Year and Test Location",
    x = "Publication Year",
    y = "Number of Studies",
    color = "Test Location"
  )

```

Interpret this graph. What are the most common test locations, and do they differ over time?

> Answer: The most common test locations are the lab, undeterminable field, natural field and artificial field. The natural field location was most common around 2010. Over time, the lab steadily increased as the most common test location. 

11. Create a bar graph of Endpoint counts. What are the two most common end points, and how are they defined? Consult the ECOTOX_CodeAppendix for more information. 

[**TIP**: Add `theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))` to the end of your plot command to rotate and align the X-axis labels...]

```{r Q11}

ggplot(Neonics, aes(x = Endpoint)) +
  geom_bar() +
  labs(
    title = "Counts of Endpoints in Neonic Studies",
    x = "Endpoint",
    y = "Number of Studies"
  )+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))

```

> Answer: The two most common endpoints are LOEL and NOEL. According to the Ecotox Code Appendix, LOEL stands for "Lowest-observable-effect-level" and is used for terrestrial areas. NOEL stands for "No-observable-effect-level" and is used for terrestrial areas. 

## Explore your data (Litter)

12. Determine the class of collectDate. Is it a date? If not, change to a date and confirm the new class of the variable. Using the `unique` function, determine which dates litter was sampled in August 2018.

```{r Q12}

class(Litter$collectDate)
#This is currently a factor class. 


Litter$collectDate <- as.Date(Litter$collectDate, format = "%Y-%m-%d")
august2018_dates <- unique(Litter$collectDate[format(Litter$collectDate, "%Y-%m") == "2018-08"])
august2018_dates

#The sampling dates of August 2018 are 8/2/2018 and 8/30/2018. 
```

13. Using the `unique` function, list the different plotIDs sampled at Niwot Ridge. How is the information obtained from `unique` different from that obtained from `summary`?

```{r Q13}

unique_plotIDs <- unique(Litter$plotID[Litter$siteID == "NIWO"])
unique_plotIDs

```

> Answer: The information obtained from the 'unique' function gives you all of the exact values within the dataset that you are searching for. The 'summary' function would give you a summary of the values instead of distinct values. 

14. Create a bar graph of functionalGroup counts. This shows you what type of litter is collected at the Niwot Ridge sites. Notice that litter types are fairly equally distributed across the Niwot Ridge sites.

```{r Q14}

niwot_data <- Litter[Litter$siteID == "NIWO", ]
ggplot(niwot_data, aes(x = functionalGroup)) +
  geom_bar() +
  labs(
    title = "Counts of Functional Groups at Niwot Ridge",
    x = "Litter Type (Functional Group)",
    y = "Number of Samples"
  )
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))

```

15. Using `geom_boxplot` and `geom_violin`, create a boxplot and a violin plot of dryMass by functionalGroup.

```{r Q15}

ggplot(Litter, aes(x = functionalGroup, y = dryMass)) +
  geom_boxplot() +
  labs(
    title = "Boxplot of Dry Mass by Functional Group",
    x = "Functional Group",
    y = "Dry Mass (g)"
  ) +
  theme(axis.text.x = element_text(angle = 45, vjust = 0.5, hjust = 1))

ggplot(Litter, aes(x = functionalGroup, y = dryMass)) +
  geom_violin() +
  labs(
    title = "Violin Plot of Dry Mass by Functional Group",
    x = "Functional Group",
    y = "Dry Mass (g)"
  ) +
  theme(axis.text.x = element_text(angle = 45, vjust = 0.5, hjust = 1))

```

Why is the boxplot a more effective visualization option than the violin plot in this case?

> Answer: The boxplot gives a more clear visualization of the dry mass values of the functional groups. It also includes information on the median, quartiles, and outliers, which is helpful when interpreting the data. 

What type(s) of litter tend to have the highest biomass at these sites?

> Answer:Needles, twigs/branches, and mixed groups of litter tend to have the highest biomasses. 
